\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-16 NAR_DRAFT.tex
fixing a bunch of inline citations and figure labels

\subsection{Real Data}
We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer \cite{Griffith2015,Shi2018}.
We downloaded the normal and primary tumor bam files generated for the whole genome sequences in \citet{Griffith2015} from dbGaP accession number phs000159.v8.p4.
\citet{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
The whole exome sequencing bam files generated by \citet{Shi2018} were downloaded from NCBI SRA no. SRP070662.
\citet{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
We called variants using Mutect 1.1.7 with the same parameters as described above, except that both of these data sets were originally aligned to GRCH37, and we used that reference here.

@@

	We use the Integrated Calibration Index to quantify the difference between MuTect and \batcave \cite{Austin2019}.
	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and \batcave (Figure \ref{NAR-wgs_fig} \textbf{e} and \textbf{f} and Table \ref{table:01}).
	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average distance between the loess fit and a diagonal line.

@@

\subsection{Real tumor data.}
We utilized two real data sets.
In each, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
These are among a very small number of tumors which have been sequenced at high depth, and for which variant validation was undertaken.
\citet{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{E} and Table \ref{table:01}).

\citet{Shi2018} performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$, and validated all variants identified by three different variant calling pipelines.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).

@@

Deep sequencing is necessary to measure quantities such as intra-tumor heterogeneity \cite{Shi2018}, sub-clonal presence of driver and resistance alleles \cite{Griffith2015}, and the mutational landscape of cancer in general \cite{Zehir2017}.  
As \citet{Griffith2015} show, whole genome sequencing at approx. 350X average depth requires combining calls from many variant calling algorithms, and then validating these calls via ultra-deep targeted sequencing. 
We believe that much of the need for multiple variant callers could be alleviated by improvements to the statistical model used to evaluate the evidence presented by the sequenced reads.

>>>>

\subsection{Real Data}
We analyzed two real data sets, one from an acute myeloid leukemia, and one from a multi-region sequencing experiment in breast cancer \cite{Griffith2015,Shi2018}.
We downloaded the normal and primary tumor bam files generated for the whole genome sequences in Griffith et al., 2015~\citep{Griffith2015} from dbGaP accession number phs000159.v8.p4.
Griffith et al., 2015~\citep{Griffith2015} generated a platinum set of variant calls for this primary tumor, which we obtained from supplementary materials, and used for our true positive dataset.
The whole exome sequencing bam files generated by Shi et al., 2015~\citep{Shi2018} were downloaded from NCBI SRA no. SRP070662.
Shi et al., 2018~\citep{Shi2018} generated a gold set of variant calls for each tumor region sequenced, which we obtained from supplementary materials and used for our true positive dataset.
We called variants using Mutect 1.1.7 with the same parameters as described above, except that both of these data sets were originally aligned to GRCH37, and we used that reference here.

@@

	We use the Integrated Calibration Index to quantify the difference between MuTect and \batcave \cite{Austin2019}.
	Briefly, a loess-smoothed regression is fit by regressing the binary (T/F) variant classification against the reported posterior probability of a true variant for both MuTect and \batcave (Figure~\ref{NAR-wgs_fig}\textbf{A}--\textbf{B} and Table~\ref{table:01}).
	Plotting the smoothed regression predictions against the actual predictions would result in a diagonal line, and the Integrated Calibration Index is the average distance between the loess fit and a diagonal line.

@@

\subsection{Real tumor data.}
We utilized two real data sets.
In each, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
These are among a very small number of tumors which have been sequenced at high depth, and for which variant validation was undertaken.
Griffith et al., 2015~\citep{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{E} and Table \ref{table:01}).

Shi et al., 2018~\citep{Shi2018} performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$, and validated all variants identified by three different variant calling pipelines.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).

@@

Deep sequencing is necessary to measure quantities such as intra-tumor heterogeneity \citep{Shi2018}, sub-clonal presence of driver and resistance alleles \cite{Griffith2015}, and the mutational landscape of cancer in general \cite{Zehir2017}.  
As Griffith et al., 2015~\citep{Griffith2015} show, whole genome sequencing at approx. 350X average depth requires combining calls from many variant calling algorithms, and then validating these calls via ultra-deep targeted sequencing. 
We believe that much of the need for multiple variant callers could be alleviated by improvements to the statistical model used to evaluate the evidence presented by the sequenced reads.
